Performance of prediction models for opportunities for improvement in trauma care across patient cohorts

### 2024-01-08 Seminarieversion.Rmd
upload utkast slutversion

---
title: "Performance of prediction models for opportunities for improvment in trauma care across patient cohorts"
author: "Anna Thoss"
output:
  pdf_document: default
  html_document: default
  word_document: default
bibliography: bibliography.bib
csl: vancouver.csl
---

```{r setup, include=FALSE, echo=FALSE}
library(knitr)
library(kableExtra)
library(flextable)

knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, comment = NA)

set_flextable_defaults(
  font.size = 8,
  theme_fun = theme_vanilla,
  padding.top = 1,
  padding.bottom = 1,
  padding.left = 1,
  padding.right = 1
)

output.format <- knitr::opts_knit$get('rmarkdown.pandoc.to')

source("./main.R")
```

```{=html}
<!-- This is an R Markdown document. Markdown is a simple formatting
syntax for authoring HTML, PDF, and MS Word documents. For more
details on using R Markdown see http://rmarkdown.rstudio.com -->
```
```{=html}
<!-- This is a comment and will not be present in the compiled
document. Try "knitting" this document by clicking the knit button up
to the left, or by running `rmarkdown::render("manuscript.Rmd")` in
the console and you should only see headings -->
```
```{=html}
<!-- Structure and most comments are from the Strengthening the
reporting of observational studies in epidemiology (STROBE) statement,
see https://www.strobe-statement.org/. For more explanations see the
paper Strengthening the Reporting of Observational Studies in
Epidemiology (STROBE):explanation and elaboration.  -->
```
```{=html}
<!-- Indicate the study’s design with a commonly used term in the
title, for example: "Associations between resuscitation interventions
and opportunities for improvement in adult trauma patients: A cohort
study" -->
```
# Abstract

```{=html}
<!-- Provide in the abstract an informative and balanced summary of
what was done and what was found. Not more than 300 words. -->
```
## Background

Trauma is among the leading causes of mortality and morbidity worldwide. The selective review of patient's mortality and morbidity cases is used to improve quality of care through identification of opportunities for improvement (OFI). The aim of this study was to compare how machine learning models perform the prediction of OFI across different trauma cohorts.

## Methods

## Results

## Conclusion

@@

![](assets/OFI-flow.png) Figure illustrating process of mortality and morbidity conference at Karolinska University Hospital. <!--Note to self: Uppdatera!-->

@@

### Missing data

<!-- Osäker vad jag ska skriva på den punkten -->

@@

<!-- skapa tabell för cohorter -->

@@

All statistical analyses were conducted using R. To extract probabilities for each patient case in the dataset, a 10-fold cross-validation strategy was employed, during which predictions for the holdout fold were recorded for each data point. We assessed performance within predefined cohorts and compared it across the different cohorts in terms of discrimination. This comparison was based on the area under the receiver operating characteristic curve (AUC). To estimate the 95% confidence interval for AUC performance, we employed bootstrapping on the extracted probabilities, repeating the process 1000 times. Additionally, an Analysis of Variance (ANOVA) test was conducted to compare differences between cohorts. A significance level of 5% was employed

@@

### Study design

<!-- Present key elements of study design -->

### Setting

```{=html}
<!-- Describe the setting, locations, and relevant dates, including
periods of recruitment, exposure, follow-up, and data collection -->
```
### Participants

```{=html}
<!-- Cohort study: Give the eligibility criteria, and the sources and
methods of selection of participants. Describe methods of
follow-up. For matched studies, give matching criteria and number of
exposed and unexposed -->
```
```{=html}
<!-- Case-control study: Give the eligibility criteria, and the
sources and + methods of case ascertainment and control
selection. Give the rationale for the choice of cases and
controls. For matched studies, give matching criteria and the number
of controls per case -->
```
```{=html}
<!-- Cross-sectional study: Give the eligibility criteria, and the
sources and methods of selection of participants -->
```
## Variables and data sources/measurements

```{=html}
<!-- Clearly define all outcomes, exposures, predictors, potential
confounders, and effect modifiers. Give diagnostic criteria, if
applicable. For each variable of interest, give sources of data and
details of methods of assessment (measurement). Describe comparability
of assessment methods if there is more than one group -->
```
## Bias

<!-- Describe any efforts to address potential sources of bias -->

## Study size

<!-- Explain how the study size was arrived at -->

## Quantitative variables

```{=html}
<!-- Explain how quantitative variables were handled in the
analyses. If applicable, describe which groupings were chosen and why
-->
```
## Statistical methods

```{=html}
<!--

(a) Describe all statistical methods, including those used to control
for confounding

(b) Describe any methods used to examine subgroups and interactions

(c) Explain how missing data were addressed

(d)

Cohort study: If applicable, explain how loss to follow-up was addressed

Case-control study:If applicable, explain how matching of cases and
controls was addressed

Cross-sectional study: If applicable, describe analytical methods
taking account of sampling strategy

(e) Describe any sensitivity analyses
-->
```
# Results

@@

```{r tab1, echo=FALSE, fig.cap="Selected sample characteristics", fig.align='center' }
if (output.format == "latex") {
  tab1 <- t1kable(create.tableone(dataset), "latex", longtable = T, booktabs = T) %>%
    kable_styling(latex_options = c("repeat_header", "condensed"), repeat_header_text = "(cont.)", font_size = 8) %>%
    kableExtra::footnote(general = "\\\\textit{Definition of  abbreviations:} OFI = Opportunity for Improvement; ED = Emergency Department; GCS = Glascow Coma Scale.\\\\newline Time to first CT and Time to definitive treatment: Measured in minutes from arrival at the hospital.", general_title = "",threeparttable =  TRUE, escape = FALSE)
} else if (output.format == "docx") {
    tab1 <- t1flex(create.tableone(dataset)) %>%
      add_footer_lines("OFI = Opportunity for Improvement; ED = Emergency Department; GCS = Glascow Coma Scale.") %>%
      autofit()
}

tab1
```

Karolinska trauma registry contains 14022 patients, of which 8309 are reviewed for OFI, either through selection by individual review or application of audit filters, as well as an unknown process (n=2). Through this, 7571 patients were selected for the morbidity conference, and 727 patients for the mortality conference.

@@

```{r main.result, echo=FALSE}
main.results.table <- flextable(cohort.performance.results) %>%
  set_caption("Area under the curve (AUC) across clinical cohorts")
```

We compared the different cohorts with AUC as our performance metric. (See table 2) Isolated severe TBI had the highest AUC at 0.811 (0.807-0.814), followed by "Other cohort" (AUC 0.761 0.76-0.762). The model that performed the lowest in the cohort "Blunt multisystem with TBI" (AUC 0.696, 0.692-0.701) Figure 2 shows details of the receiver operating curve. The P-value between the different cohorts was 1 x 10 \^-6.

```{r roc.curves, out.width="80%",fig.cap="ROC Curves. LÄGG CAPTION HÄR", echo=FALSE, fig.align='center'}
knitr::include_graphics("figures/roc.pdf", error = FALSE)
```

```{=html}
<!--

Participants
------------

(a) Report numbers of individuals at each stage of study—eg numbers
potentially eligible, examined for eligibility, confirmed eligible,
included in the study, completing follow-up, and analysed

(b) Give reasons for non-participation at each stage

(c) Consider use of a flow diagram

Descriptive data
----------------

(a) Give characteristics of study participants (eg demographic,
clinical, social) and information on exposures and potential
confounders

(b) Indicate number of participants with missing data for each
variable of interest

(c) Cohort study — Summarise follow-up time (eg, average and total
amount)

Outcome data
------------

Cohort study — Report numbers of outcome events or summary measures
over time

Case-control study — Report numbers in each exposure category, or
summary measures of exposure

Cross-sectional study — Report numbers of outcome events or summary
measures

Main results
------------

(a) Give unadjusted estimates and, if applicable, confounder-adjusted
estimates and their precision (eg, 95% confidence interval). Make
clear which confounders were adjusted for and why they were included

(b) Report category boundaries when continuous variables were
categorized

(c) If relevant, consider translating estimates of relative risk into
absolute risk for a meaningful time period

Other analyses 
--------------

Report other analyses done—eg analyses of subgroups and interactions,
and sensitivity analyses 
-->
```
You can include code in this document like this:

```{r main, echo=FALSE}
source("main.R") ## This "imports" the main script file of your project and run any code in it
```

You can also embed plots:

```{r plot, echo=FALSE}
plot(pressure)
```

You can also mix text and code, so called inline code, like this: `r 2+5`.

@@

### Key results

We set out to assess machine learning models viability as a tool to aid clinicians in their effort to improve trauma care quality, and specifically how they would perform across different trauma cohorts. We found that there is some variance in the models performance between the different cohorts, with "Blunt multisystem without TBI" performing the lowest, despite being the largest clearly defined trauma cohort in the study. The largest group "Other cohorts", which is made up of cases falling outside of the boundaries for the other formalized cohorts performs the best over all, despite possibly encompassing the largest heterogeneity.

A previous study on the same dataset has found that audit filters, with an AUC of 0.624, are outperformed by our models when the dataset is not categorised by cohorts.[@Attergrim2023] It is unclear how audit filters perform in this context across cohorts as this was outside of the scope the study. Interestingly, the model performed better on the "Isolated severe TBI" cohort in terms of AUC then the best model in the aforementioned study (AUC 0.811 vs 0.789). A 2020 study found similarly good performance of machine learning models in TBI patients looking to predict poor outcomes. [@Matsuo2020] This may suggest that some models inclination to perform well in prediction of traumatic brain injury, whereas outcomes linked to other trauma cohorts may be harder to predict.

@@

Secondly, our learners are dependent on the current selection process for OFI, as this makes out the training data and thus makes out the groundwork for what is considered OFI in our dataset. As a consequence, the models are prone to some of the same limitations as the current system of audit filters. The review process is dependent both the individual review of specialized nurses, along with the current set of audit filters, which may not be the optimal parameters for demeriting OFI. This introduces an inherent bias. The audit filter system favour some types of error, while possibly missing others. At Karolinska University Hospital, the current system is more apt at identifying severe clinical outcomes, such as preventable death, or patient cases with more interventions or longer hospital stays, and flagging them as possible OFI. Higher ISS scores and "Days in hospital" have been found to be predictors of high importance when identifying OFI [@Albaaj2023]. This in turn introduces possible selection bias, giving a skewed view of errors within the trauma care, and as such not a holistic view of the trauma care quality.

@@

```{=html}
<!--

Key results
-----------
Summarise key results with reference to study objectives

Limitations
-----------

Discuss limitations of the study, taking into account sources of
potential bias or imprecision.  Discuss both direction and magnitude
of any potential bias

Interpretation
--------------

Give a cautious overall interpretation of results considering
objectives, limitations, multiplicity of analyses, results from
similar studies, and other relevant evidence

Generalisability
----------------

Discuss the generalisability (external validity) of the study results

-->
```
# Conclusion

>>>>

---
title: "Performance of prediction models for opportunities for improvement in trauma care across patient cohorts"
author: "Anna Thoss"
output:
  word_document: default
  html_document: default
  pdf_document: default
bibliography: bibliography.bib
csl: vancouver.csl
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

@@

Figure illustrating process of mortality and morbidity conference at Karolinska University Hospital.

@@

@@

@@

All statistical analyses were conducted using R. To extract probabilities for each patient case in the dataset, a 10-fold cross-validation strategy was employed, during which predictions for the holdout fold were recorded for each data point. We assessed performance within predefined cohorts and compared it across the different cohorts in terms of discrimination. This comparison was based on the area under the receiver operating characteristic curve (AUC). To estimate the 95% confidence interval for AUC performance, we employed bootstrapping on the extracted probabilities, repeating the process 1000 times. Additionally, an Analysis of Variance (ANOVA) test was conducted to compare differences between cohorts. A significance level of 5% was employed.

@@

# Results

@@

The Karolinska trauma registry includes 14022 patients, of which 8309 are reviewed for OFI, either through selection by individual review or application of audit filters, as well as an unknown process (n=2). Through this, 7571 patients were selected for the morbidity conference, and 727 patients for the mortality conference.

@@

We compared the performance of the ML model in the different cohorts with AUC as our performance metric (See table 2). The ML model performed best in the cohort of patients with isolated severe TBI, in which it had an AUC of 0.811 (0.807-0.814), followed by "Other cohort" (AUC 0.761 0.76-0.762). The ML model performed worst in the cohort of patients with blunt multisystem trauma with TBI (AUC 0.696, 0.692-0.701). Figure 2 shows details of the receiver operating curves. The P-value between the different cohorts was 1 x 10 \^-6, meaning that there were significant differences in AUC between cohorts.

@@

We set out to assess the performance of a model based on the machine learner XGBoost's viability as a tool to aid clinicians in their effort to improve trauma care quality, and specifically how they would perform across different trauma cohorts. We found that there is some variation in the model's performance between the different cohorts, with "Blunt multisystem without TBI" performing the lowest, despite being the largest clearly defined trauma cohort in the study. The largest group "Other cohorts", which is made up of cases falling outside of the boundaries for the other formalized cohorts performs the best over all, despite possibly encompassing the largest heterogeneity.

A previous study on the same dataset has found that audit filters, with an AUC of 0.624, are outperformed by ML models when the dataset is not categorised by cohorts.[@Attergrim2023] It is unclear how audit filters perform in this context across cohorts as this was outside of the scope the study. Interestingly, the model performed better on the "Isolated severe TBI" cohort in terms of AUC then the best model in the aforementioned study (AUC 0.811 vs 0.789). A 2020 study found similarly good performance of machine learning models in TBI patients looking to predict poor outcomes. [@Matsuo2020] This may suggest that some models inclination to perform well in prediction of traumatic brain injury, whereas outcomes linked to other trauma cohorts may be harder to predict.

@@

Secondly, our learner is dependent on the current selection process for OFI, as this makes out the training data and thus makes out the groundwork for what is considered OFI in our dataset. As a consequence, the models are prone to some of the same limitations as the current system of audit filters. The review process is dependent both the individual review of specialized nurses, along with the current set of audit filters, which may not be the optimal parameters for determining OFI. This introduces an inherent bias. The audit filter system favour some types of error, while possibly missing others. At Karolinska University Hospital, the current system is more apt at identifying severe clinical outcomes, such as preventable death, or patient cases with more interventions or longer hospital stays, and flagging them as possible OFI. Higher ISS scores and "Days in hospital" have been found to be predictors of high importance when identifying OFI [@Albaaj2023]. This in turn introduces possible selection bias, giving a skewed view of errors within the trauma care, and as such not a holistic view of the trauma care quality.

@@

# Conclusion
